HO unification from object language to meta language

### 2024-04-01 paper/main.tex
shorten 7

\subsection{Limitations of by this basic scheme}
\label{sec:basic-comp-limitations}

\begin{gather}
\lambda x y.F \appsep y \appsep x = \lambda x y.x \label{eq:unif-eta1}\\
\lambda x. f \appsep (F \appsep x) \appsep x = G \label{eq:unif-eta2}
% \lambda x. f \appsep (F \appsep x) \appsep x = f \appsep (\lambda y.y) \label{eq:unif-eta2}
\end{gather}
Note that here $F$ is used with different arities, moreover
in the second problem the left hand side happens to be an
eta expansion (of $f (\lambda y.y)$) only after we discover (at run time)
that $F = \lambda x\lambda y.y$ (i.e. that $F$ discards the $x$ argument).
Both problems are addressed in the next section.

@@

In \cref{sec:basic-comp-limitations}, we have given two unification problems to
be run one after the other. In the following table we present the entry problem,
its compiled version, with the corresponding list of mapping and links.
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{X,E,1},
    {Y,F,0},
    {X,C,2}}}
  {{{,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {,A,\lambda x.B_{x}},
    {x,B_{x},\lambda y.C_{y x}}}}

\marginpar{aggiustare i nomi delle variabili nel testo}
We see that the \elpiIn{maybe-eta} as detected $\lambda xy.F \appsep y \appsep
x$ and $\lambda x.f\appsep (F \appsep x) \appsep x$ and replaced them with
respectively the \Ho vars $X$ and $Z$. $X$ is linked with $\lambda x.Y_x$, $Y$
has arity $1$ and is $\eta$-linked with $\lambda y.H\appsep y \appsep x$ and $Z$
is linked to the term $\lambda x.f \appsep G_{x}~x$. However, the mapping
returned by the compilation, does not breaks \cref{inv:map-bijective}: the \Fo{}
variable $F$ is mapped to two different \Ho variables. To address this problem
and enforce \cref{inv:map-bijective}, we clean the mapping with a second phase
after the compilation. This phase is called \elpiIn{map-deduplication}.

Before formally defining this procedure, we need to define some auxiliary
relations. Let \mapStore be the list of mapping and $\langle m_1, m_2 \rangle
\in \mapStore$ such that the arity of the \Ho variable in $m_1$ is smaller than
the one in $m_2$. Let $X$ (resp. $Y$) the \Ho variable of $m_1$ (resp. $m_2$)
and $n = ar(m_1) - ar(m_2)$. We also let $A^i$ be a fresh \Ho variable. We
define the \elpiIn{make-eta-link} relation taking two mappings $\langle m_1, m_2
\rangle$ and returning the following list of link \linketa: $\forall i \in [1..n]$,
{
\def\scopeA{\ensuremath{x_1\dots x_{i-1}}}
\def\scopeB{\ensuremath{x_1\dots x_{i}}}
%
$$ 
  \left\{ 
    \begin{array}{ll}
      \linketaM{\phantom{\scopeA}}{X}      {\lambda x.  A^1_x}         & if~ i = 1 \\
      \linketaM{\scopeA}{A^{i-1}_{\scopeA}}{\lambda x_i.A^i_{\scopeB}} & if~ 1 < i < n\\
      \linketaM{\scopeA}{A^{i-1}_{\scopeA}}{\lambda x_i.Y  _{\scopeB}} & if~ i = n
    \end{array} 
  \right. 
$$
}
\noindent
More concretely, we are saying that for any two mappings, we build as many
\linketa as the difference of the arities between the two mappings. This links
are constructed in such a way that the \Ho variable $v$ with lowest arity is
linked to a fresh variable eta-expended variable $A^1$ having the scope of $v$.
This variable $A^1$ is then linked to a an $\eta$-expanded fresh variable $A^2$
with same scope of $A^1$ and so on. The last link is built between the $A^{n-1}$
(where $n$ is the difference of arities between the two mappings) and the \Ho
variable $u$ with higher arity in the two mappings being considered.

\begin{definition}[map-deduplication]
  Forall mappings $\langle m_1, m_2 \rangle \in \mapStore$, sharing the same
  \Fo{} variable, the list of \linketa $L$ is created thanks to
  \elpiIn{make-eta-link}$m_1~m_2~L$ and is added to \linkStore. Then $m_1$ is
  removed from \mapStore.
\end{definition}

If we take back the example give at the beginning of this section, we can
deduplicate $\mapping{F}{G}{1}, \mapping{F}{H}{2}$ by removing the first mapping
and adding the auxiliary \linketa: \linketaM{x}{G_{x}}{\lambda y.H_{x y}}.

The complete problem to run for resolution is now:
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{Y,F,0},
    {X,C,2}}}
  {{{x,E_{x},\lambda y.C_{x y}},
    {,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {,A,\lambda x.B_{x}},
    {x,B_{x},\lambda y.C_{y x}}}}

After unification of the two terms, $X$ is assigned to $\lambda x.\lambda y.x$.
This assignment makes $l_2$ to progress since the \lhs is materialized and by
unification, between $X$ and $\lambda x.Y_{x}$, $Y_x$ is instantiate to $\lambda
y.x$. Once $Y_x$ is instantiated, $l_1$ can progress, and set $H_{xy}$ to $x$.
After all these progresses, $l_1$ and $l_2$ are remove from \linkStore and the
\elpiIn{progress} fixpoint terminates. Next, the second unification problem is
run, and $Z$ is set to $f\appsep(\lambda x.x)$. This unification wakes up $l_3$
and since $Z$ starts with the \elpiIn{app} node, the $\eta$-expanded version of
$Z$ is unified with $\lambda x.f \appsep G_{x}~x$ and $G_x$ is set to $x$.
As last step, the last link is progressed and the final \Ho substitution is
$\{X _{} \mapsto \lambda x.\lambda y.x, Y _{x} \mapsto \lambda y.x, 
G _{yx} \mapsto y, Z _{} \mapsto f~\lambda x.x, 
H _{x} \mapsto \lambda y.y\}$.

The decompilation phase is only charged, in this example to solve the mappings,
since no suspended links remain. The only mapping in the list is
\mapping{F}{H}{2}, which will assign the $F$ variable in \Fo{} to $\lambda xy.y$

>>>>

\subsection{Limitations of by this basic scheme}
\label{sec:basic-comp-limitations}
The basic compilation scheme is not about to
deal wit the following problem:
\printAlll
  {{{\lambda x y.X \appsep y \appsep x, \lambda x y.x},{\lambda x. f \appsep (X \appsep x) \appsep x,G}}}
  {{}}
  {{}}
  {{}}
\marginpar{TODO: remove empty T}
% \begin{gather}
% \lambda x y.F \appsep y \appsep x = \lambda x y.x \label{eq:unif-eta1}\\
% \lambda x. f \appsep (F \appsep x) \appsep x = G \label{eq:unif-eta2}
% % \lambda x. f \appsep (F \appsep x) \appsep x = f \appsep (\lambda y.y) \label{eq:unif-eta2}
% \end{gather}

noindent Note that here $X$ is used with different arities, moreover
in the second problem the left hand side happens to be an
eta expansion (of $f (\lambda y.y)$) only after we discover (at run time)
that $X = \lambda x\lambda y.y$ (i.e. that $X$ discards the $x$ argument).
Both problems are addressed in the next two sections.

@@

We report here the problem given in \cref{sec:basic-comp-limitations} where
$X$ is used with two different arities and the output of the compilation does
not respect\cref{inv:map-bijective} (merging the two mappings
for $s$ would break \ref{inv:uvaarity}).
In this section we explain how to replace the duplicate mapping with some
\linketa{} in order to restore the invariants.
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{X,E,1},
    {Y,F,0},
    {X,C,2}}}
  {{{,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {,A,\lambda x.B_{x}},
    {x,B_{x},\lambda y.C_{y x}}}}

We see that the \elpiIn{maybe-eta} as identified $\lambda xy.X \appsep y \appsep
x$ and $\lambda x.f\appsep (X \appsep x) \appsep x$ and the compiler has
replaced them with $A$ and $D$ respectively.
% $A$ is linked with $\lambda x.B_x$, $B$
% has arity $1$ and is $\eta$-linked with $\lambda y.C\appsep y \appsep x$ and $D$
% is linked to the term $\lambda x.f \appsep E_{x}~x$. 
However, the mapping \mapStore breaks \cref{inv:map-bijective}: the \Fo{}
variable $X$ is mapped to two different \Ho variables. To address this problem
we adjust the compiler's output with a \elpiIn{map-deduplication} procedure.

\newcommand{\alignarity}{\emph{align-arity}}
\begin{definition}[\alignarity] Given two mappings
  $m_1 : X \mapsto A^m$ and
  $m_2 : X \mapsto C^n$ where $m < n$ and $d = n - m$,
  $\alignarity{}~ m_1 ~ m_2$ generates the following $d$ links, one
  for each $i$ such that $0 \leq i < d$,
\[%
  \linketaM{x_0 \ldots x_{m+i}}
           {B^i_{x_0 \ldots x_{m+i}}}
           {\lambda x_{m+i+1}.B^{i+1}_{x_0 \ldots x_{m+i+1}}}
\]%
where $B^i$ is a fresh variable of arity $m+i$, and $B^0 = A$ as well as $B^d = C$.
\end{definition}

The intuition is that we $\eta$-expand the occurrence of the variable
with lower arity to match the higher arity. Since each \linketa can
add exactly one lambda, we need as many links as the difference between the
two arities.

\newcommand{\mapdeduplication}{\emph{map-deduplication}}

\begin{definition}[\mapdeduplication]
  Forall mappings $m_1, m_2\in \mapStore$ such that 
  $m_1 : X \mapsto A^m$ and
  $m_2 : X \mapsto C^n$ and $m < n$
  we remove $m_1$ from \mapStore and
  add to \linkStore the result of $\alignarity{}~m_1~m_2$.
\end{definition}

If we look back the example give at the beginning of this section, we can
deduplicate $\mapping{X}{E}{1}, \mapping{X}{C}{2}$ by removing the first mapping
and adding the auxiliary \linketa: \linketaM{x}{E_{x}}{\lambda y.C_{x y}}.
After deduplication the compiler output is as follows:
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{Y,F,0},
    {X,C,2}}}
  {{{x,E_{x},\lambda y.C_{x y}},
    {,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {,A,\lambda x.B_{x}},
    {x,B_{x},\lambda y.C_{y x}}}}
\marginpar{TODO: non so se serva spiegare l'esecuzione (in commento)}
% After unification of the two terms, $X$ is assigned to $\lambda x.\lambda y.x$.
% This assignment makes $l_2$ to progress since the \lhs is materialized and by
% unification, between $X$ and $\lambda x.Y_{x}$, $Y_x$ is instantiate to $\lambda
% y.x$. Once $Y_x$ is instantiated, $l_1$ can progress, and set $H_{xy}$ to $x$.
% After all these progresses, $l_1$ and $l_2$ are remove from \linkStore and the
% \elpiIn{progress} fixpoint terminates. Next, the second unification problem is
% run, and $Z$ is set to $f\appsep(\lambda x.x)$. This unification wakes up $l_3$
% and since $Z$ starts with the \elpiIn{app} node, the $\eta$-expanded version of
% $Z$ is unified with $\lambda x.f \appsep G_{x}~x$ and $G_x$ is set to $x$.
% As last step, the last link is progressed and the final \Ho substitution is
% $\{X _{} \mapsto \lambda x.\lambda y.x, Y _{x} \mapsto \lambda y.x, 
% G _{yx} \mapsto y, Z _{} \mapsto f~\lambda x.x, 
% H _{x} \mapsto \lambda y.y\}$.
% The decompilation phase is only charged, in this example to solve the mappings,
% since no suspended links remain. The only mapping in the list is
% \mapping{F}{H}{2}, which will assign the $F$ variable in \Fo{} to $\lambda xy.y$
